Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-15 manuscript/main.tex
phylo ctmc done

%\newcommand{\TTTheta}{\boldsymbol{\Theta}}
%\newcommand{\tr}{\mbox{tr}}
%\newcommand{\X}{\mathbf{X}}
%\renewcommand{\u}{\mathbf{u}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\DD}{\mathbf{D}}
\newcommand{\MM}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\RR}{\mathbf{R}}
%\newcommand{\haar}{\mathcal{H}}
%\newcommand{\orthog}{\mathcal{O}}
%\newcommand{\SSigma}{\boldsymbol{\Sigma}}
\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Llambda}{\boldsymbol{\Lambda}}
\newcommand{\vecc}{\mbox{vec}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\BB}{\mathbf{B}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\GG}{\mathbf{G}}
\newcommand{\mom}{\boldsymbol{\xi}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\renewcommand{\P}{\mathbf{P}}

@@

\subsection{Surrogate trajectory HMC}

Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\Y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\Y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

@@

\subsection{Phylogeographic CTMC model}

Start with a (possibly unknown) rooted and bifurcating phylogenetic tree $\phylogeny$ consisting of $N$ leaf nodes that correspond to observed biological specimens and $N-1$ internal nodes that correspond to unobserved ancestors.  The tree also contains $2N-2$ branches of length $t_v$ connecting each child node $v$ to its parent $u$.  
 Given $\phylogeny$, we model the evolution of $S$ characters along each branch of the tree according to a CTMC model with $S\times S$ generator matrix $\QQ$ and stationary distribution $\ppi=(\pi_1,\dots,\pi_S)=\lim_{t\rightarrow \infty}e^{t\QQ}\widetilde{\ppi}$, for $\widetilde{\ppi}$ for any arbitrary probability vector.   Examples of characters are the $S=4$ nucleotides within a set of aligned genome sequences  \citep{jukes1969evolution} and the set of $S=15$ geographic regions visited by an influenza subtype \citep{lemey2014unifying}.   Depending on the given problem, we may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$.   In the former case, one may augment the model with a rate scalar $\gamma$ that modulates the expected number of substitutions across all branches, and the finite-time transition probability matrix along branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.

>>>>

%\newcommand{\TTTheta}{\boldsymbol{\Theta}}
%\newcommand{\tr}{\mbox{tr}}
%\newcommand{\X}{\mathbf{X}}
%\renewcommand{\u}{\mathbf{u}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\DD}{\mathbf{D}}
\newcommand{\MM}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\RR}{\mathbf{R}}
%\newcommand{\haar}{\mathcal{H}}
%\newcommand{\orthog}{\mathcal{O}}
%\newcommand{\SSigma}{\boldsymbol{\Sigma}}
\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Llambda}{\boldsymbol{\Lambda}}
\newcommand{\vecc}{\mbox{vec}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\BB}{\mathbf{B}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\GG}{\mathbf{G}}
\newcommand{\mom}{\boldsymbol{\xi}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\renewcommand{\P}{\mathbf{P}}
\newcommand{\p}{\mathbf{p}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\B}{\mathbf{B}}

@@

\subsection{Surrogate trajectory HMC}\label{sec:surrHmc}

Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid,neal2011mcmc} is an advanced MCMC procedure that uses numerical approximations to Hamiltonian trajectories to generate Metropolis-type \citep{metropolis1953equation} proposals far away from the current Markov chain state.  On the one hand, this approach to proposal generation helps reduce autocorrelation between chain states and is particularly helpful within higher-dimensional state spaces.  On the other hand, numerical integration of Hamilton's equations requires repeated evaluation of the Hamiltonian potential energy's gradient, and these repeated floating-point operations may become computationally burdensome.   Using HMC for Bayesian inference, the potential energy takes the form $u(\ttheta)=-\log \pi(\ttheta)$, where $\pi(\cdot)$ is the posterior density function $\pi(\ttheta):=p(\ttheta|\y)$ of parameter $\ttheta\in\mathbb{R}^D$ given observed data $\y$.  In this context, a single iteration of the the St\"{o}rmer-Verlet (velocity Verlet) or leapfrog integrator \citep{leimkuhler2004simulating}, beginning at time $\tau$ and with $\epsilon>0$ small, takes the form
\begin{align}\label{eq:leap}
	\mom\left(\tau + \frac{\epsilon}{2}\right) &= \mom(\tau) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau))  \\  \nonumber
	\ttheta(\tau+\epsilon) &= \ttheta(\tau) + \epsilon \, \MM^{-1}\mom(\tau+\frac{\epsilon}{2}) \\ \nonumber
	\mom(\tau + \epsilon) &= \mom\left(\tau+\frac{\epsilon}{2}\right) + \frac{\epsilon}{2} \nabla \log \pi(\ttheta(\tau+\epsilon))  \, ,
\end{align}
where $\mom \sim \mbox{N}_D(\Zero,\MM)$ is the Hamiltonian system's Gaussian-distributed momentum variable.  
Different strategies aim to speed up the leapfrog integrator's many log-posterior gradient evaluations $\nabla \log \pi (\ttheta)$ in different ways, as these numerical routines often represent the algorithm's computational bottleneck.  In model-specific contexts, \citet{holbrook2021massive,holbrook2022bayesian,holbrook2022viral} develop parallelization strategies, and \citet{ji2020gradients} develop dynamic programming techniques, to accelerate $\nabla \log \pi (\ttheta)$.

@@

\subsection{Phylogenetic CTMC model}

Start with a (possibly unknown) rooted and bifurcating phylogenetic tree $\phylogeny$ consisting of $N$ leaf nodes that correspond to observed biological specimens and $N-1$ internal nodes that correspond to unobserved ancestors.  The tree also contains $2N-2$ branches of length $t_v$ connecting each child node $v$ to its parent $u$.  
 Given $\phylogeny$, we model the evolution of $S$ characters along each branch of the tree according to a CTMC model with $S\times S$ generator matrix $\QQ$ and stationary distribution $\ppi=(\pi_1,\dots,\pi_S)=\lim_{t\rightarrow \infty}e^{t\QQ}\widetilde{\ppi}$, for $\widetilde{\ppi}$ for any arbitrary probability vector.   Examples of characters are the $S=4$ nucleotides within a set of aligned genome sequences  \citep{jukes1969evolution} and the set of $S=15$ geographic regions visited by an influenza subtype \citep{lemey2014unifying}.   We may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$ depending on the given problem.   In the former case, one may augment the model with a rate scalar $\gamma$ that modulates the expected number of substitutions across all branches, and the finite-time transition probability matrix along branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.  In the following, we further posit $\QQ=\QQ(\ttheta)$ for $\ttheta$ a vector of parameters.

 Let data $\y$ be the $N$-vector with elements $y_n\in \{1,\dots,S\}$, the observed states of each observed biological specimen $n \in \{1,\dots,N\}$.  At any node $v$, the likelihood takes the form \citep{ji2020gradients}
 \begin{align}
p(\y|\ttheta) = \p_v^T \q_v \, ,
 \end{align}
where $\p_v$ and $\q_v$ are the post-order and pre-order partial likelihood vectors, respectively.  The former describes the probability of the observed states for all observed specimens (i.e., leaf nodes) that descend from node $v$, conditioned on the state at node $v$. The latter describes analogous probabilities for all observed specimens not descending from node $v$.  If we suppose that node $u$ gives rise to two child nodes, $v$ and $w$, then 
\begin{align}\label{eq:recursions}
	\p_u = \P_v \p_v \circ \P_w \p_w\, ,  \quad \q_v= \P_v^T \left( \q_u \circ  \P_w \p_w\right) \, .
\end{align}
Using the chain rule and the fact that $\p_v$ does not depend on $\P_v$, one may write the likelihood's derivative with respect to a single parameter $\theta_d$ thus:
\begin{align}\label{eq:phyloGrad}
	\frac{\partial}{\partial \theta_d}  p(\y|\ttheta)  &=   \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial \P_v}{\partial \theta_d} \right) = \sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \frac{\partial e^{\gamma t_v \QQ}}{\partial \theta_d} \right)  \\
	&=\sum_{v=1}^{2N-2} \mbox{tr} \left( \frac{\partial p(\y|\ttheta)}{\partial \P_v} \sum_{s=1}^S \sum_{s'=1}^S \frac{\partial e^{\gamma t_v \QQ}}{\partial q_{ss'}} \frac{\partial q_{ss'}}{\partial \theta_d} \right)  \, .
\end{align}
Whereas the recursions of \eqref{eq:recursions} facilitate fast likelihood computation, inferring $\ttheta$ using gradient-based approaches such as HMC (Section \ref{sec:surrHmc}) requires a large number of repeated evaluations of the matrix exponential derivative.  These computations become particularly onerous when one opts for a gold-standard mixed-effect model \textcolor{red}{(MAGEE)} and specifies
\begin{align}
\log q_{ss'} = b_{ss'} + \epsilon_{ss'} \,, \quad s\neq s'\, .
\end{align} 
Here, the fixed-effects $b_{ss'}$ are elements of some non-random matrix $\B$, and the random-effects $\epsilon_{ss'}$ are mutually independent \emph{a priori} and inferred as model parameters.  The dimension $D$ of $\ttheta$ in this model is $\mathcal{O}(S^2)$, so the usual $\mathcal{O}(DNS^5)$ cost of the matrix expoential derivative becomes a massive $\mathcal{O}(NS^7)$.  In this context, \textcolor{red}{MAGEE} show that the first-approximation \eqref{eq:firstOrder} achieves a considerable speedup over the exact derivative, while facilitating high-quality proposals in the context of surrogate HMC.  In the following section, we use this method to analyze the global spread of SARS-CoV-2 and show that the first-order approximation maintains its performance for an even higher-dimensional problem than previously considered.
